Fast, Accurate, and Scalable Genomic Structural Variation Detection on the Cloud: MapReduce based {\em Cloudbreak}

### 2013-07-03 manuscript/nm.tex
update abstract, editing

 \begin{abstract}
The detection of genomic structural variations (SV) remains one of the the most difficult challenges in analyzing high-throughput sequencing data due to the uncertainty inherent in short-read sequencing data and the  scale of the data sets. The growing size and number of sequencing data sets have rendered SV detection, and high throughput sequencing analysis in general, a bona fide big data problem and led to intense interest in distributing computation to cloud or commodity servers. Meanwhile, MapReduce is becoming a standard framework for distributing processing across compute clusters and has convincingly demonstrated its effectiveness in offering scalable solutions to big data problems. We describe a novel conceptual framework for SV detection algorithms in MapReduce based on computing local features along the genome. In this framework, we develop and evaluate a distributed deletion and insertion detection algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome. On simulated and real data sets, our approach achieves performances meeting or exceeding those of a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes. In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp. In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples. Finally, our software offers dramatically faster runtimes than existing tools, given a sufficiently sized compute cluster, and also includes an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, making cluster computing accessible to users who may not have the resources to maintain a MapReduce capable distributed compute server. Our implementation and source code are available at \url{http://github.com/cwhelan/cloudbreak}.

@@

We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive. This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage.

To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same sample. These included data from the Human Genome Structural Variation Project reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507. We merged any overlapping calls of the same type into the region spanned by their unions. It should be noted that the 1000 Genomes call set was partially produced using DELLY and BreakDancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor. We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.

@@

Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard set for both deletions and insertions. All algorithms show far less specificity for the gold standard set than they did for the true variants in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample. For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels. Cloudbreak has slightly lower accuracy for insertions than other tools, although it can identify the most variants at higher levels of sensitivity. Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, BreakDancer, even when BreakDancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).

Given the high number of false positives produced by all tools at maximum sensitivity indicated by the ROC curves, we decided to characterize the predictions made by each tool at more stringent thresholds. We examined the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X. For insertions, again, we were forced to use the thresholds that gave maximum sensitivity for each tool due to the high observed FDR rates in the simulated data. The precision and recall at these thresholds with respect to the gold standard set, as well as the performance of each algorithm at predicting variants of each size class at those thresholds, is shown in Table \ref{NA18507DeletionAndInsertionPreds}. For deletions, Cloudbreak has the greatest sensitivity of any tool at these thresholds, identifying the most variants in each size class. Pindel exhibits the highest precision with respect to the gold standard set. For insertions, Pindel again has the highest precision at maximum sensitivity, although according the ROC curve it is possible to choose a threshold for Cloudbreak with higher precision and recall. At maximum sensitivity, Cloudbreak identifies 120 more insertions from the gold standard set than Pindel, giving it by far the highest recall.

@@

Because Cloudbreak explicitly models zygosity in its feature generation algorithm, it is possible to predict the genotypes of identified variants. We tested this on both the simulated and NA18507 data sets. Because we simulated a diploid genome, we had genotypes for the simulated data readily available. For the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}. Cloudbreak was able to achieve 92.7\% and 95.9\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets, respectively. Supplementary Table \ref{SdeletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier. None of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertions that meet our size threshold that also had genotyping information. Of the 123 insertions detected by Cloudbreak on the simulated data set, 43 were heterozygous. Cloudbreak correctly classified 78 of the 80 homozygous insertions and 31 of the 43 heterozygous insertions discovered, for an overall accuracy rate of 88.6\%.

>>>>

 \begin{abstract}

The detection of genomic structural variations (SV) remains one of the the most difficult challenges in analyzing high-throughput sequencing data due to the uncertainty inherent in mapping short sequencing reads and the scale of the data. The growing size and number of sequencing data sets have rendered SV detection, and high throughput sequencing analysis in general, a bona fide big data problem and led to intense interest in distributing computation to clusters of commodity servers or on-demand cloud infrastructure. Meanwhile, MapReduce is becoming a standard framework for distributed computing and has convincingly demonstrated its effectiveness in offering scalable solutions to big data problems. We describe a novel conceptual framework for SV detection algorithms in MapReduce based on computing local features along the genome. In this framework, we develop and evaluate a distributed deletion and insertion detection algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome. On simulated and real data sets, our approach achieves performances meeting or exceeding those of a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes, while providing dramatically shorter runtimes and the ability to readily scale up in data volume on large clusters. In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp and can accurately genotype heterozygous and homozygous deletions from diploid samples. Finally, our software distribution includes an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, enabling on-demand cluster computing. Our implementation and source code are available at \url{http://github.com/cwhelan/cloudbreak}.

@@

We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive. This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage. To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same sample. These included data from the Human Genome Structural Variation Project reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507. We merged any overlapping calls of the same type into the region spanned by their unions. It should be noted that the 1000 Genomes call set was partially produced using DELLY and BreakDancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor. We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.

@@

Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard set for both deletions and insertions. All algorithms show far less specificity for the gold standard set than they did for the true variants in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample. For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels. Cloudbreak has comparable accuracy for insertions to other tools, and can identify the most variants at higher levels of sensitivity. Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, BreakDancer, even when BreakDancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).

Given the high number of false positives produced by all tools at maximum sensitivity, we decided to characterize the predictions made at more stringent thresholds. We examined the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X. For insertions, we were again forced to use the maximum sensitivity thresholds for each tool due to the high observed FDRs in the simulated data. The precision and recall at these thresholds with respect to the gold standard set, as well as the performance of each algorithm at predicting variants of each size class, is shown in Table \ref{NA18507DeletionAndInsertionPreds}. For deletions, Cloudbreak has the greatest sensitivity of any tool, identifying the most variants in each size class. Pindel exhibits the highest precision with respect to the gold standard set. For insertions, Pindel again has the highest precision at maximum sensitivity, although according the ROC curve it is possible to choose a threshold for Cloudbreak with higher precision and recall. At maximum sensitivity, Cloudbreak identifies 120 more insertions from the gold standard set than Pindel, giving it by far the highest recall.

@@

Because Cloudbreak explicitly models zygosity in its feature generation algorithm, it can predict the genotypes of identified variants. We tested this on both the simulated and NA18507 data sets. Because we simulated a diploid genome, genotypes for the simulated data were readily available. For the real dataset, we considered the deletions from the 1000 Genomes Project data set, which had been genotyped using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}. Cloudbreak was able to achieve 92.7\% and 95.9\% accuracy in predicting the genotype of the deletions it detected at our 10\% FDR threshold in the simulated and real data sets, respectively. Supplementary Table \ref{SdeletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier. None of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertions that met our size threshold and also had genotyping information. Of the 123 insertions detected by Cloudbreak on the simulated data set, 43 were heterozygous. Cloudbreak correctly classified 78 of the 80 homozygous insertions and 31 of the 43 heterozygous insertions, for an overall accuracy of 88.6\%.
